---
title: "Storm Analysis"
author: "Nick Daughenbaugh"
date: "August 6, 2020"
output:
  pdf_document: default
  html_document: default
---


## 1: Assignment

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

Questions
Your data analysis must address the following questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?


## 2: Data Analysis

### 2.1: Downloading & Reading-in Data

Download the raw data file and extract the data into a dataframe.Then convert to a data.table
```{r DataLoading}
library("data.table")
library("ggplot2")
library("tinytex")
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, destfile = paste0("C:/Users/nrdau/Documents/R/data/StormData.csv.bz2"))
stormDF <- read.csv("C:/Users/nrdau/Documents/R/data/StormData.csv.bz2")
# Converting data.frame to data.table
stormDT <- as.data.table(stormDF)
```

### 2.2: Expoling Data Fields

```{r ColumnNames}
colnames(stormDT)
```

### 2.3: Data Subsetting

Remove the fields we don't need. 
```{r DataSubsetting, results="hide"}
# Finding fields to (not) remove
Remove <- colnames(stormDT[, !c("EVTYPE"
  , "FATALITIES"
  , "INJURIES"
  , "PROPDMG"
  , "PROPDMGEXP"
  , "CROPDMG"
  , "CROPDMGEXP")])
# Removing columns
stormDT[, c(Remove) := NULL]
# Only use data where fatalities or injuries occurred.  
stormDT <- stormDT[(EVTYPE != "?" & 
             (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), c(
               "EVTYPE",
               "FATALITIES",
               "INJURIES",
               "PROPDMG",
               "PROPDMGEXP",
               "CROPDMG",
               "CROPDMGEXP") ]
```

### 2.4: Converting Exponent Prefixes

Covert the PROPDMGEXP and CROPDMGEXP fields to something useful.

```{r CorrectingExponents, results="hide"}
# Change all damage exponents to uppercase.
EXPcols <- c("PROPDMGEXP", "CROPDMGEXP")
stormDT[,  (EXPcols) := c(lapply(.SD, toupper)), .SDcols = EXPcols]
# Map property damage alphanumeric exponents to numeric values.
propDmgKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)
# Convert crop damage prefix exponents to numeric values.
cropDmgKey <-  c("\"\"" = 10^0,
                "?" = 10^0, 
                "0" = 10^0,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)
stormDT[, PROPDMGEXP := propDmgKey[as.character(stormDT[,PROPDMGEXP])]]
stormDT[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]
stormDT[, CROPDMGEXP := cropDmgKey[as.character(stormDT[,CROPDMGEXP])] ]
stormDT[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]
```

### 2.5: Economic Damage

```{r EconomicCostColumns}
stormDT <- stormDT[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, propCost = PROPDMG * PROPDMGEXP, CROPDMG, CROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP)]
```

### 2.6: Total Property and Crop Cost

```{r TotalPropertyCropCost}
totalCostDT <- stormDT[, .(propCost = sum(propCost), cropCost = sum(cropCost), Total_Cost = sum(propCost) + sum(cropCost)), by = .(EVTYPE)]
totalCostDT <- totalCostDT[order(-Total_Cost), ]
totalCostDT <- totalCostDT[1:10, ]
head(totalCostDT, 5)
```

### 2.7: Total Fatalities and Injuries

```{r TotalFatalitiesInjuriesCalc}
totalInjuriesDT <- stormDT[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]
totalInjuriesDT <- totalInjuriesDT[order(-FATALITIES), ]
totalInjuriesDT <- totalInjuriesDT[1:10, ]
head(totalInjuriesDT, 5)
```

## 3: Conclusion

### 3.1: The Most Harmful Events to Population

Melting data.table so that it is easier to put in bar graph format 
```{r HealthResults}
harmful <- melt(totalInjuriesDT, id.vars="EVTYPE", variable.name = "Harm")
head(harmful, 5)
```

```{r healthChart}
# Create chart
healthChart <- ggplot(harmful, aes(x=reorder(EVTYPE, -value), y=value))
# Plot data as bar chart
healthChart = healthChart + geom_bar(stat="identity", aes(fill=Harm), position="dodge")
# Format y-axis scale and set y-axis label
healthChart = healthChart + ylab("Frequency Count") 
# Set x-axis label
healthChart = healthChart + xlab("Event Type") 
# Rotate x-axis tick labels 
healthChart = healthChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
healthChart = healthChart + ggtitle("Top 10 Deadliest US Storm Events") + theme(plot.title = element_text(hjust = 0.5))
healthChart
```

### 3.2: The Most Harmful Events to Economic Output

Melting data.table so that it is easier to put in bar graph format 
```{r EconConsequences}
econ_output <- melt(totalCostDT, id.vars="EVTYPE", variable.name = "Damage_Type")
head(econ_output, 5)
```

```{r econChart}
# Create chart
econChart <- ggplot(econ_output, aes(x=reorder(EVTYPE, -value), y=value))
# Plot data as bar chart
econChart = econChart + geom_bar(stat="identity", aes(fill=Damage_Type), position="dodge")
# Format y-axis scale and set y-axis label
econChart = econChart + ylab("Cost (dollars)") 
# Set x-axis label
econChart = econChart + xlab("Event Type") 
# Rotate x-axis tick labels 
econChart = econChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
econChart = econChart + ggtitle("Top 10 US Storm Events Damaging Economic Output") + theme(plot.title = element_text(hjust = 0.5))
econChart
```
